Topological Equivariant Artist Model for Visualization Library Architecture

### 2025-03-05 paper/main.tex
start of generalizing presheaf

\begin{definition}\cite{bradleyWhatFunctorDefinitions,bradleyTopologyCategoricalApproach2020} A \textbf{functor} is a map $F: \mathcal{C} \rightarrow \mathcal{D}$, which means it is a function between objects $F: \textbf{ob}(\mathcal{C}) \mapsto \textbf{ob}(\mathcal{D})$ and that for every morphism $f \in Hom(C_1, C_2)$  there is a corresponding function $F: Hom(C1, C2) \mapsto Hom(F(C_1), F( C_2))$.
A \textbf{functor} must satisfy the properties
\begin{itemize}
  \item \textit{identity}: $F(id_{C}(C)) = id_{D}(F(C))$
  \item \textit{composition}: $F(g)\circ F(f) = F(g\circ f)$ for any composable morphisms $C_{1}\xrightarrow{f} C_2$, $C_2 \xrightarrow{g} C_3$
\end{itemize}
$F(C) \in \textbf{ob}(\mathcal{D})$ denotes the object to which an object $C$ is mapped, and $F(f) \in Hom(F_(C_1), F_(C_2))$ denotes the morphism that $f$ is mapped to.
\end{definition}
Modeling the data container as a functor allows us state that, just like a functor, the container is a map between index space objects and sets of data records that preserve morphisms between index space objects and data records.
\begin{equation}
  \label{eq:atct:sheaf:functor}
  \sheafc_{\dbasec, \dtotalc}: \opensetc \rightarrow \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}}
\end{equation}
A common way of encapsulating a map from a topological space to a category of sets is as a presheaf
\begin{definition}\label{def:atct:presheaf}
  A \textbf{presheaf} $F:\mathcal{C}^{op} \rightarrow \setb$ is a contravariant functor from an object in an arbitrary category to an object in the category \setb\cite{nlab:presheaf, spanier1989algebraic}.
\end{definition}
A functor is contravariant when the morphisms between the input objects go in the opposite direction from the morphisms between the output objects. The presheaf is contravariant because the inclusion morphisms between input object $\iota: \opensetc_1 \rightarrow \opensetc_2$
are defined such that they correspond to the partial ordering $\openset_1 \subseteq \openset_2$, but the restriction morphisms $\iota^*$ between the sets of sections $\iota^*: \cgamma{\opensetc_2}{\dtotalc\restriction_{\opensetc_2}} \rightarrow \cgamma{\opensetc_1}{\dtotalc\restriction_{\openset_c1}}$ restricts the larger set to the smaller one such that all functions that are continuous over a space must be continuous over a subspace $\Gamma_2 \subseteq \Gamma_1$, where $\Gamma_{i}\coloneqq\Gamma(\openset_{i}, \dtotal\restriction_{\openset_{i}})$.

@@

\subsection{Equivariance}
\label{sec:artist:equiv}
As introduced in \autoref{sec:related-work:equivariance}, data and the corresponding visual encoding are expected to have compatible structure such that actions (\autoref{def:related-work:action}) on data and graphic are symmetric. We generalize from binary operations to a family of actions $\dfunc \in \dfuncset$ on the sheaf $\sheaf_{\dbase, \dtotal}$ because that allows for expanding the set of allowable transformations on the data beyond a single operator. We describe the changes on the graphic side as changes in measurements \measure\ of visual components. A superset of the Bertin visual variables \cite{\cite{bertinIIPropertiesGraphic2011}, measurable components are scaler or vector components of the rendered graphic that can be quantified, such as the color, position, shape, texture, or rotation angle of the graphic. For example, a measurement of a scatter marker could be its color (e.g. red) or its x position (e.g. 5).

@@

As mentioned in \autoref{sec:artist:homeomorphic}, we construct the data base space as a deformation retraction of the graphic space. On simple way of doing so is to construct the graphic base space as a constant multiple of the base space such that
\begin{equation}
  \underbrace{\dbasec\times[0,1]^{n}}_{\gbasec} \textcolor{functor}{\xmapsto{\hspace{1em}\vindexc\hspace{1em}}} \dbasec
\end{equation}
where n is a thickening of the graphic base space $\gbase$ to account for the dimensionality of the output space
\begin{equation*}
  n = \begin{cases}
    dim(\gbase) - dim(\dbase) & dim(\dbase)<dim(\gbase)\\
  0 & otherwise
  \end{cases}
\end{equation*}
because otherwise the data dimensionality $\dbase$ may be too small for a graphic representation. For example, as shown in \autoref{fig:construction:xi}, a line is 1D but is a 2D glyph on a screen; therefore the graphic space $\gbase$ is constructed by multiplying the base space $\dbase$ with an interval $[0,1]$.

>>>>

\begin{definition}\cite{bradleyWhatFunctorDefinitions,bradleyTopologyCategoricalApproach2020} A \textbf{functor} is a map $F: \mathcal{C} \rightarrow \mathcal{D}$, which means it is a function between objects $F: \textbf{ob}(\mathcal{C}) \mapsto \textbf{ob}(\mathcal{D})$ and that for every morphism $f \in Hom(C_1, C_2)$  there is a corresponding function $F: Hom(C1, C2) \mapsto Hom(F(C_1), F( C_2))$.
A \textbf{functor} must satisfy the properties
\begin{itemize}
  \item \textit{identity}: $F(id_{C}(C)) = id_{D}(F(C))$
  \item \textit{composition}: $F(g)\circ F(f) = F(g\circ f)$ for any composable morphisms $C_{1}\xrightarrow{f} C_2$, $C_2 \xrightarrow{g} C_3$
\end{itemize}
$F(C) \in \textbf{ob}(\mathcal{D})$ denotes the object to which an object $C$ is mapped, and $F(f) \in Hom(F_(C_1), F_(C_2))$ denotes the morphism that $f$ is mapped to.
\end{definition}
Modeling the data container as a functor allows us state that, just like a functor, the container is a map between index space objects and sets of data records that preserve morphisms between index space objects and data records.
\begin{equation}
  \label{eq:atct:sheaf:functor}
  \sheafc_{\dbasec, \dtotalc}: \opensetc \rightarrow \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}}
\end{equation}
A common way of encapsulating a map from a topological space to a category of sets is as a presheaf
\begin{definition}\label{def:atct:presheaf}
  A \textbf{presheaf} $F:\mathcal{C}^{op} \rightarrow \setb$ is a contravariant functor from an object in an arbitrary category to an object in the category \setb\cite{spanier1989algebraic}.
\end{definition}
A functor is contravariant when the morphisms between the input objects go in the opposite direction from the morphisms between the output objects. The presheaf is contravariant when for every arbitrary morphism between input base spaces $\dfunch: \openset_1 \rightarrow \openset_2$ there exists a corresponding pullback function between the sets of sections $\dfunchpull:  \cgamma{\opensetc_2}{\dtotalc\restriction_{\opensetc_2}} \rightarrow \cgamma{\opensetc_1}{\dtotalc\restriction_{\openset_c1}}$. The job

%% these two paragraphs cen be condensed and moved below the figure as an example
The presheaf is contravariant because the morphisms between input object $\iota: \opensetc_1 \rightarrow \opensetc_2$
are defined such that they correspond to the partial ordering $\openset_1 \subseteq \openset_2$, but the restriction morphisms $\iota^*$ between the sets of sections $\iota^*: \cgamma{\opensetc_2}{\dtotalc\restriction_{\opensetc_2}} \rightarrow \cgamma{\opensetc_1}{\dtotalc\restriction_{\openset_c1}}$ restricts the larger set to the smaller one such that all functions that are continuous over a space must be continuous over a subspace $\Gamma_2 \subseteq \Gamma_1$, where $\Gamma_{i}\coloneqq\Gamma(\openset_{i}, \dtotal\restriction_{\openset_{i}})$.

@@

\subsection{Equivariance}
\label{sec:artist:equiv}
As introduced in \autoref{sec:related-work:equivariance}, data and the corresponding visual encoding are expected to have compatible structure such that actions (\autoref{def:related-work:action}) on data and graphic are symmetric. We generalize from binary operations to a family of actions $\dfunc \in \dfuncset$ on the sheaf $\sheaf_{\dbase, \dtotal}$ because that allows for expanding the set of allowable transformations on the data beyond a single operator. We describe the changes on the graphic side as changes in \textcolor{monoid}{measurements} \measurec\ of visual components. A superset of the Bertin visual variables \cite{bertinIIPropertiesGraphic2011}, measurable components are scaler or vector components of the rendered graphic that can be quantified, such as the color, position, shape, texture, or rotation angle of the graphic. For example, a measurement of a scatter marker could be its color (e.g. red) or its x position (e.g. 5).

@@

\label{sec:construction:xi}
As mentioned in \autoref{sec:artist:homeomorphic}, we construct the data base space as a deformation retraction of the graphic space. On simple way of doing so is to construct the graphic base space as a constant multiple of the base space such that
\begin{equation}
  \underbrace{\dbasec\times[0,1]^{n}}_{\gbasec} \textcolor{functor}{\xmapsto{\hspace{1em}\vindexc\hspace{1em}}} \dbasec
\end{equation}
where n is a thickening of the graphic base space $\gbase$ to account for the dimensionality of the output space
\begin{equation*}
  n = \begin{cases}
    dim(\gbase) - dim(\dbase) & dim(\dbase)<dim(\gbase)\\
  0 & otherwise
  \end{cases}
\end{equation*}
because otherwise the data dimensionality $\dbase$ may be too small for a graphic representation. For example, as shown in \autoref{fig:construction:xi}, a line is 1D but is a 2D glyph on a screen; therefore the graphic space $\gbase$ is constructed by multiplying the base space $\dbase$ with an interval $[0,1]$.
